Semistochastic importance sampling of second-quantized operators in determinant space

### 2021-06-07 notes/talk/talk.tex
Final version?

\documentclass[amsmath]{beamer}
\setbeamertemplate{navigation symbols}{} % Turn off navigation symbols
\usepackage{beamerthemeshadow}
\usepackage{dcolumn}               % Align table columns on decimal point
\usepackage{framed}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{multicol}
\usepackage[export]{adjustbox}
\usepackage{biblatex}
\usepackage{pdfpages}
\bibliography{hci.bib}
\usefonttheme[onlymath]{serif}

@@

\begin{frame}{Determinant-space algorithms}
	Slater determinants, second-quantized operators in this basis
\end{frame}

@@

\section{Algorithm}
\subsection{Deterministic component}
\begin{frame}{Deterministic component}
	\Wider{
		\begin{itemize}
			\item  \textbf{Setup:}
			\begin{itemize}
				\item  For each pair of orbitals $\left\{p,q\right\}$, store a list of %triplets
				$\left\{r,s,\left|H\left(rs\leftarrow pq\right)\right|\right\}$,
				sorted by $\left|H\left(rs\leftarrow pq\right)\right|$ in decreasing order
				\item Also compute (for sampling algorithm):
				\begin{eqnarray}
				\sum_{rs\in\textrm{remaining}} |H(rs\leftarrow pq)|
				\end{eqnarray}
			\end{itemize}
			{\color{white}  skip line}
			\\
			\uncover<2->{
				\item \textbf{Finding important connected determinants} $D_j$ (for which $\left|H_{ji}\right|>\epsilon$): %for at least one $i\in \mathcal{V}$)
				\\{\color{white}  skip line}
				\\
			}
			\begin{itemize}
				\uncover<3->{
					%\item  \textbf{Generate only those double excitations that exceed $\epsilon$.}
					%\\
					\item For each pair of occupied orbitals $\left\{p,q\right\}$, look up the stored list
					of triplets $\left\{r,s,\left|H\left(rs\leftarrow pq\right)\right|\right\}$ and iterate
					through the list until a triplet is reached for which $\left|H\left(rs\leftarrow pq\right)\right|<\epsilon$
				}
				\\{\color{white}  skip line}
				\\
			\end{itemize}
		\end{itemize}
	}
\end{frame}

@@

\subsection{Stochastic component}
\begin{frame}{Discrete sampling algorithms}
	\begin{itemize}
		\item \textbf{Alias sampling:}
		\begin{enumerate}
			\item Sample an element with uniform probability
			\item If a low-probability element was selected, it has some probability of `aliasing' to another, high-probability element
		\end{enumerate}
	    Setup time: $\mathcal{O}(N)$, Time per sample: $\mathcal{O}(1)$
		\item \textbf{Cumulative distribution function (CDF) searching:}
		\begin{enumerate}
			\item Sample a real number $r$ uniformly between 0 and 1
			\item Binary-search the CDF for $r$ (i.e., find the smallest index $i$ for which $r<\textrm{CDF}_i$)
		\end{enumerate}	
		Setup time: $\mathcal{O}(N)$, Time per sample: $\mathcal{O}(\log N)$
	\end{itemize}
Usually, Alias sampling is preferred because it has lower sampling complexity, but CDF searching has a unique use case that we will need...
\end{frame}

@@

\begin{frame}{Importance-sampling excitations}
	\begin{itemize}
	\item Recall that we have already described how to treat the large-magnitude ($|H_{ji}|\ge \epsilon$) excitations deterministically
	\item Now, we want to ``importance-sample'' the remaining small-magnitude ($|H_{ji}|<\epsilon$) excitations, i.e.:
	\begin{eqnarray}
		P(H_{ji}) \begin{cases}
			\propto  f(|H_{ji}|),&\textrm{if $|H_{ji}| < \epsilon$};\\
			=  0, &\textrm{if $|H_{ji}|\ge\epsilon$}. % (because we already treated these deterministically)}.
		\end{cases}
	\end{eqnarray}
	\end{itemize}
	Examples of importance-sampling distributions include: \begin{eqnarray}f(|H_{ji}|) = |H_{ji}|,\quad \quad f(|H_{ji}|) = |H_{ji}|^2\end{eqnarray}
\end{frame}

>>>>

\documentclass[amsmath]{beamer}
\setbeamertemplate{navigation symbols}{} % Turn off navigation symbols
\usepackage{beamerthemeshadow}
\usepackage{dcolumn}               % Align table columns on decimal point
\usepackage{framed}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{multicol}
\usepackage[export]{adjustbox}
\usepackage{biblatex}
\usepackage{pdfpages}
\bibliography{hci.bib}
\usefonttheme[onlymath]{serif}
\beamerdefaultoverlayspecification{<+->}

@@

%\begin{frame}{Determinant-space algorithms}
%	Slater determinants, second-quantized operators in this basis
%\end{frame}

@@

\section{Algorithm}
\subsection{Deterministic component}
\begin{frame}{Deterministic component}
	\Wider{
		\begin{itemize}
			\item  \textbf{Setup:}
			\begin{itemize}
				\item  For each pair of orbitals $\left\{p,q\right\}$, store a list of %triplets
				$\left\{r,s,\left|H\left(rs\leftarrow pq\right)\right|\right\}$,
				sorted by $\left|H\left(rs\leftarrow pq\right)\right|$ in decreasing order
				\linebreak
				\item Also compute (for sampling algorithm):
				\begin{eqnarray}
				\sum_{rs\in\textrm{remaining}} |H(rs\leftarrow pq)|
				\end{eqnarray}
			\end{itemize}
			{\color{white}  skip line}
			\\
			\uncover<2->{
				\item \textbf{Finding important connected determinants} $D_j$ (for which $\left|H_{ji}\right|>\epsilon$): %for at least one $i\in \mathcal{V}$)
				\\{\color{white}  skip line}
				\\
			}
			\begin{itemize}
				\uncover<3->{
					%\item  \textbf{Generate only those double excitations that exceed $\epsilon$.}
					%\\
					\item For each pair of occupied orbitals $\left\{p,q\right\}$, look up the stored list
					of triplets $\left\{r,s,\left|H\left(rs\leftarrow pq\right)\right|\right\}$ and iterate
					through the list until a triplet is reached for which $\left|H\left(rs\leftarrow pq\right)\right|<\epsilon$
				}
				\\{\color{white}  skip line}
				\\
			\end{itemize}
		\end{itemize}
	}
\end{frame}

@@

\subsection{Stochastic component}
\begin{frame}{Discrete sampling algorithms}
	\begin{itemize}
		\item \textbf{Alias sampling:}
		\begin{enumerate}
			\item Sample an element with uniform probability
			\item If a low-probability element was selected, it has some probability of `aliasing' to another, high-probability element
		\end{enumerate}
	    \begin{itemize}
	    	\item Setup time: $\mathcal{O}(N)$, Time per sample: $\mathcal{O}(1)$
    	\end{itemize}
		\item \textbf{Cumulative distribution function (CDF) searching:}
		\begin{enumerate}
			\item Sample a real number $r$ uniformly between 0 and 1
			\item Binary-search the CDF for $r$ (i.e., find the smallest index $i$ for which $r<\textrm{CDF}_i$)
		\end{enumerate}	
		\begin{itemize}
			\item Setup time: $\mathcal{O}(N)$, Time per sample: $\mathcal{O}(\log N)$
		\end{itemize}
		\item Usually, Alias sampling is preferred because it has lower sampling complexity, but CDF searching has a unique use case that we will need...
	\end{itemize}
\end{frame}

@@

\begin{frame}{Importance-sampling excitations}
	\begin{itemize}
	\item Recall that we have already described how to treat the large-magnitude ($|H_{ji}|\ge \epsilon$) excitations deterministically
	\item Now, we want to ``importance-sample'' the remaining small-magnitude ($|H_{ji}|<\epsilon$) excitations, i.e.:
	\begin{eqnarray}
		P(H_{ji}) \begin{cases}
			\propto  f(|H_{ji}|),&\textrm{if $|H_{ji}| < \epsilon$};\\
			=  0, &\textrm{if $|H_{ji}|\ge\epsilon$}. % (because we already treated these deterministically)}.
		\end{cases}
	\end{eqnarray}
	\item Examples of importance-sampling distributions include: \begin{eqnarray}f(|H_{ji}|) = |H_{ji}|,\quad \quad f(|H_{ji}|) = |H_{ji}|^2\end{eqnarray}
	\end{itemize}
\end{frame}
